---
title: "NLP with R - Part 2: sentiment analysis"
author: Sonia Mazzi<br> [ONS - Data Science Campus](https://datasciencecampus.ons.gov.uk)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_caption: false
    fig_width: 11
    fig_height: 6
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true 
    theme: cosmo
    df_print: paged
---
 ```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("pics/dsclogo.png"),
               alt = "logo", 
               style = "position: absolute; width: 180px; top:10px; right:10px; padding: 10px")
```


# Setting up

**Required libraries**

We will first load the libraries we will be using in what follows.

If you don't have the libraries installed you may need to execute this first

```{r eval=FALSE}
install.packages(c("readr", "dplyr", "tidyr", "stringr", "ggplot2", "ggrepel", "kableExtra", "formattable", "gridExtra", "tidytext", "textdata", "memery", "magick", "circlize", "radarchart", "igraph", "gggraph", "widyr"))
```

Now you can load the libraries

```{r message=F, warning=F}
library(readr)# read text files
library(dplyr) #data manipulation
library(tidyr) #data wrangling
library(stringr) #manipulate strings
library(ggplot2) #visualizations
library(ggrepel)
library(kableExtra)
library(formattable)
library(gridExtra) #viewing multiple ggplots in a grid
library(tidytext) #text mining
library(textdata) #needed for loading sentiment dictionaries
library(memery) #add a background image to plots - if you have a Mac you need to install XQuartz in you computer from XQuartz.org - for more fun see https://cran.r-project.org/web/packages/memery/vignettes/memery.html
library(magick)
library(circlize)
library(radarchart)
library(igraph)
library(ggraph)
library(widyr)
```

<br>

**ggplot pre-set theme**

```{r}
theme_prince <- function(aticks = element_blank(),
                         pgminor = element_blank(),
                         lt = element_blank(),
                         lp = "none")
{
  theme(plot.title = element_text(hjust = 0.5), #Center the title
        axis.ticks = aticks, #Set axis ticks to on or off
        panel.grid.minor = pgminor, #Turn the minor grid lines on or off
        legend.title = lt, #Turn the legend title on or off
        legend.position = lp) #Turn the legend on or off
}
```

**Table styling pre-set**

```{r}
#this function is to print a table using kable and kableExtra
my_kable_styling <- function(dat, caption) {
  kable(dat, "html", escape = FALSE, caption = caption) %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "bordered"),
                full_width = FALSE)
}
```

**Define some colors to use throughout**

```{r}
my_colors <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#D55E00", "#D65E00")
```


# Introduction to Part 2

Every data science project needs to have a set of questions to explore. Here are a few to keep in mind as you work through this tutorial: is it possible to write a program to determine the mood expressed in lyrics? Are predefined lexicons sufficient? How much data preparation is necessary? Can you link your results to real-life events? Does sentiment change over time? Are hit songs more positive or negative than uncharted songs? What words stand out in the lyrics during the year Prince was said to predict 9/11? Did he predict his own death?



On April 21, 1985, Prince recorded "Sometimes It Snows In April", <https://www.youtube.com/watch?v=ikZgBhSMSUM>.
He died exactly 31 years later on April 21, 2016. Did prince predict his own death?
Let us carry out a sentiment analysis of the song.

First we load the data we saved in Part 1 consisting on the tidy version of prince_new.csv, i.e the tokenized version without 1) stop words, 2) undesirable words (as defined in Part 1), and 3) 1-3 character words, except some pronouns. 

```{r warning=F, message=F}
prince_tidy_filtered <- read_csv("data/prince_tidy_filtered.csv")
glimpse(prince_tidy_filtered)
```


Let us add a sentiment to the tokens using the NRC lexicon

```{r}
prince_tidy_filtered_nrc <- inner_join(prince_tidy_filtered, get_sentiments("nrc") )
glimpse(prince_tidy_filtered_nrc)
```

Let us visualise the sentiment contained in the song "Sometimes It Snows in April".


```{r}
prince_tidy_filtered_nrc_plot <- prince_tidy_filtered_nrc %>% 
  filter(song == "sometimes it snows in april") %>%
  group_by(sentiment) %>%
  count(word, sort = TRUE) %>%
  arrange(desc(n)) %>%
  ungroup()
```

```{r}
prince_tidy_filtered_nrc_plot %>%
  ggplot(aes(x = word, y = 1, label = word, fill = sentiment)) +  #set `y = 1` to just plot one variable and use word as the label
  geom_point(colour = "transparent") +
  geom_label_repel(force = 1, nudge_y = 0.5, direction = "y", box.padding = 0.04, segment.color = "transparent", size = 3) +
  facet_grid(~sentiment) +
  theme_prince() +
  theme(axis.text.y = element_blank(), axis.text.x = element_blank(),
        axis.title.x = element_text(size = 6),
        panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect("lightgray", fill = NA),
        strip.text.x = element_text(size = 9)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("'Sometimes it snows in April' - Sentiment visualization") + 
  coord_flip()
```

What do you think?



# Lexical diversity

In what follows we will carry out sentiment analysis of Prince's lyrics.

The method consists of using pre-defined lexicons.

The results of sentiment analysis depend on the level at which the analysis is carried out.
These levels are typically identified as document, sentence, word, etc. 
In lyrics, the level could be defined as sentiment per decade, year, chart-level, or song. 
The sentence level is not usually an option with lyrics as punctuation can detract from rhymes and patterns. 
Word level analysis exposes detailed information and can be used as foundation knowledge for more advanced practices in topic modeling.


```{r}
words_by_decade <- prince_tidy_filtered %>%
  mutate(decade = ifelse(is.na(decade),"NONE", decade)) %>% #turn NA into NONE
  group_by(decade, song) %>%
  mutate(word_count = n_distinct(word)) %>%
  select(song, Released = decade, Charted = charted, word_count) %>%
  distinct() %>% #To obtain one record per song
  ungroup()
glimpse(words_by_decade)
```

```{r}
words_by_decade %>%
  ggplot(aes(x = Released, y = word_count, colour = Released)) +
  geom_boxplot() +
  facet_grid(~Charted) +
  theme_prince() +
  labs(x = "", y = "Nr of different words of a song") +
  ggtitle("Lexical diversity by decade")
```

There is an upwards trend in the lexical diversity as the decades go by until the 1990s, when the trend more or less flattens, both for charted and uncharted songs. Note the amazing spread of lexical diversity for uncharted songs with an unknown release date.


# Chords diagrams: Charted Songs By Decade

A chord diagram allows to study flows between a set of entities. Entities (nodes) are displayed all around a circle and connected with arcs (links, chords). In R, the `circlize` package is the best option to build it.

The following graph shows the relationship between the decade a song was released and whether or not it hit the Billboard charts.
The graph is split into two categories: charted (top), and decade (bottom). The two variables are separated by wide gaps, with smaller gaps between the levels of each variable. 

```{r}
prince_data <- read_csv("data/prince_new.csv")
```


```{r}
decade_chart <-  prince_data %>%
  filter(decade != "NA") %>% #Remove songs without release dates
  count(decade, charted)  #Get SONG count per chart level per decade. Order determines top or bottom.
decade_chart
```

```{r}
circos.clear() #Very important - Reset the circular layout parameters!
grid_col <- c("1970s" = my_colors[1], "1980s" = my_colors[2], "1990s" = my_colors[3], "2000s" = my_colors[4], "2010s" = my_colors[5], "Charted" = "grey", "Uncharted" = "grey") #assign chord colors
# Set the global parameters for the circular layout. Specifically the gap size
decade_aux <- decade_chart %>% select(decade) %>% n_distinct() #nr of unique values of decade
charted_aux <- decade_chart %>% select(charted) %>% n_distinct() #nr of unique values of charted


circos.par(gap.after = c(rep(5, decade_aux - 1), 15,
                         rep(5, charted_aux - 1), 15))#gaps between levels are always 5 and between factors 15

chordDiagram(decade_chart, grid.col = grid_col, transparency = .2)
title("Relationship Between Chart and Decade")
```

The above circle graph may seem complex at first glance, but it nicely illustrates the counts of songs per decade, per chart level. We can see that Prince began his career in the 1970s with only a few releases, some of which charted. If you compare the 1980s to the 1990s, you'll find that more songs were released in the 1990s, but more songs charted in the 1980s. There were only a few commercially successful songs in the 2000s and in the 2010s there were no hit songs.

# Choosing a lexicon for sentiment analysis

In order to determine which lexicon to use for sentiment analysis of Prince's lyrics, we will look at the match ratio of words that are common to both the lexicon and the lyrics. As a reminder, there are 36046 total words and 7212 distinct words in `prince_tidy_filtered`.

```{r}
nrow(prince_tidy_filtered)
```

```{r}
prince_tidy_filtered %>% select(word) %>% n_distinct()
```


How many of those words are actually in the lexicons?

Use  `inner_join()` between `prince_tidy` and `new_sentiments` and then group by lexicon. The NRC lexicon has 10 different categories, and a word may appear in more than one category: that is, words can be, for example, negative and sad. We will use `n_distinct()` in `summarise()` to get the distinct word count per lexicon.

```{r}
sentiments_new <- read_csv("sentiments_new.csv")
```


```{r}
prince_tidy_filtered %>%
  mutate(words_in_lyrics = n_distinct(word)) %>%
  inner_join(sentiments_new) %>%
  group_by(lexicon, words_in_lyrics, words_in_lexicon) %>%
  summarise(lex_match_words = n_distinct(word)) %>%
  ungroup() %>%
  mutate(total_match_words = sum(lex_match_words), #Not used but good to have
         match_ratio = lex_match_words / words_in_lyrics) %>%
  select(lexicon, lex_match_words, match_ratio) %>%
  mutate(lex_match_words = color_bar("lightpink")(lex_match_words),
         lexicon = color_tile("lightgreen", "lightgreen")(lexicon)) %>%
  my_kable_styling(caption = "Lyrics words found in lexicons")
```

The NRC lexicon contains the highest number of unique lyrics words. Notice the sum of the match ratios is low. No lexicon could have all words, nor should they. Many words are considered neutral and would not have an associated sentiment. For example, "2000" is typically a neutral word, and therefore does not exist in the lexicons. However, if you remember, people predicted planes would fall out of the sky and computers would just stop working during that year. So there is an associated fear that exists in the song but is not captured in sentiment analysis using typical lexicons.

Here are a few reasons that a word may not appear in a lexicon:

*Not every word has a sentiment.

*The lexicons were created for other types of text, so not for lyrics.

*The actual form of the word may not appear. For example, "strong" may appear, but "strongly" may not. There could be more cleaning needed on the data.

<br>

<br>

Take a look at some specific words from Prince's lyrics which seem like they would have an impact on sentiment. Are they in all lexicons?

```{r}
sentiments_new %>%
  filter(word %in% c("dark", "controversy", "gangster",
                     "discouraged", "race")) %>%
  arrange(word) %>% #sort
  mutate(word = color_tile("lightblue", "lightblue")(word),
         words_in_lexicon = color_bar("lightpink")(words_in_lexicon),
         lexicon = color_tile("lightgreen", "lightgreen")(lexicon)) %>%
  my_kable_styling(caption = "Specific Words")
```

"Controversy" and "dark" appear in NRC and BING, but "gangster" only appears in BING. "Race" doesn't appear at all and is a critical topic in Prince's lyrics. 
But is it easily associated with a sentiment? Note that AFINN is much smaller and only has one of these words.

<br>

# Word Forms

Now look at a more complicated example. Sexuality is a common theme in Prince's music. 
How will sentiment analysis based on predefined lexicons be affected by different forms of a word? 
For example, here are all the references to the root word "sex" in the lyrics. Compare these to BING and NRC and see where there are matches.

```{r}
prince_data <- read_csv("data/prince_new.csv")
```

```{r}
my_word_list <- prince_data %>%
  unnest_tokens(word, lyrics) %>%
  mutate(aux = str_detect(word, "sex")) %>%  #find the substring `"sex"`
  filter(aux == "TRUE") %>% 
  count(word) %>%
  select(myword = word, n) %>% #Rename word
  arrange(desc(n))

sentiments_new %>%
  #Right join gets all words in `my_word_list` to show nulls
  right_join(my_word_list, by = c("word" = "myword")) %>%
  filter(word %in% my_word_list$myword) %>%
  mutate(word = color_tile("lightblue", "lightblue")(word),
          instances = color_tile("lightpink", "lightpink")(n),
          lexicon = color_tile("lightgreen", "lightgreen")(lexicon)) %>%
  select(-n) %>% #Remove these fields
  my_kable_styling(caption = "Dependency on Word Form")
```  
  
  Notice that Prince uses "sexy" frequently, but the word doesn't appear in this form in the lexicon NRC. The word "sex" is found in NRC but not in Bing. What if we looked at the stems or roots of words, would that help? The text being analysed could contain a past tense, a plural, or an adverb of a root word, but it may not exist in any lexicon. How do you deal with this?
  
 
 More Data Preparation

It may be the case that you need a few more data preparation steps. Here are three techniques to consider before performing sentiment analysis:

* **Stemming**: generally refers to removing suffixes from words to get the common origin

* **Lemmatization**: reducing inflected (or sometimes derived) words to their word stem, base or root form

* **Word replacement**: replace words with more frequently used synonyms

An advanced concept in sentiment analysis is that of synonym (semantically similar peer) and hypernym (a common parent) replacement. These are words that are more frequently used than the related word in the lyric, and actually do appear in a lexicon, thus giving a higher match percentage. There is not enough space in this tutorial to address additional data preparation, but it's definitely something to consider!


<br>

<br>

# Sentiment Analysis

Now that you have a foundational understanding of the data set and the lexicons, you can apply that knowledge by joining them together for analysis. 
Here are the high-level steps you'll take:

* Create lexicon-specific data sets.

* Look at polar sentiment across all songs.

* Examine sentiment change over time.

* Validate your results against specific events in Prince's life.

* Study song level sentiment.

* Review how pairs of words affect sentiment.

<br>

<br>

As we did with the poems of Plath and Rumi, we will investigate the sentiment in all of Prince's lyrics according to the lexicons.

Let us use BING for binary (positive/negative) and NRC for categorical sentiments. 
Since words can appear in multiple categories in NRC, such as Negative/Fear or Positive/Joy, we will also create a subset without the positive and negative categories to use later on.

```{r message=F}
prince_bing <- prince_tidy_filtered %>%
  inner_join(get_sentiments("bing"))

prince_nrc <- prince_tidy_filtered%>%
  inner_join(get_sentiments("nrc"))

prince_nrc_sub <- prince_tidy_filtered %>%
  inner_join(get_sentiments("nrc")) %>%
  filter(!sentiment %in% c("positive", "negative"))
```  

```{r}
prince_nrc
```
 
```{r}
prince_nrc %>%
  group_by(sentiment) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, word_count))
```
  

```{r}
nrc_plot <- prince_nrc %>%
  group_by(sentiment) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, word_count)) %>%
  #Use `fill = -word_count` to make the larger bars darker
  ggplot(aes(sentiment, word_count, fill = -word_count)) +
  geom_col(na.rm = TRUE) +
  guides(fill = FALSE) + #Turn off the legend
  theme_prince() +
  labs(x = NULL, y = "Word Count") +
  ggtitle("Prince NRC Sentiment") +
  coord_flip()

img <- "pics/prince_background.png" #Load the background image
lab <- ""  #Turn off the label
#Overlay the plot on the image and create the meme file
meme(img, lab, "pics/meme_nrc.jpg", inset = nrc_plot)
#Read the file back in and display it!
nrc_meme <- image_read("pics/meme_nrc.jpg") # image_read() is from the magick package
plot(nrc_meme)
```
 
 Using the NRC lexicon we find that a positive sentiment prevails in Prince's lyrics. 
 But are all words with a sentiment of disgust or anger also in the negative category as well? It may be worth checking out.
 
 <br>
 
**EXERCISE.** Now take a look at BING overall sentiment. Of the 1150 distinct words from Prince's lyrics that appear in the BING lexicon, how many are positive and how many are negative? Display the results in a meme as above using "pics/prince_background4.jpg" as background picture. In `ggplot()`, map `fill` to sentiment. Comment on the results.

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>



```{r}
bing_plot <- prince_bing %>%
  group_by(sentiment) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, word_count)) %>%
  ggplot(aes(sentiment, word_count, fill = sentiment)) +
  geom_col() +
  guides(fill = FALSE) +
  theme_prince() +
  labs(x = NULL, y = "Word Count") +
  ggtitle("Prince Bing Sentiment") +
  coord_flip()

img1 <- "pics/prince_background4.jpg"
lab1 <- ""
meme(img1, lab1, "pics/meme_bing.jpg", inset = bing_plot)
x <- image_read("pics/meme_bing.jpg")
plot(x)
```

<br>

# Polarity and positivity indeces

Let us use the BING lexicon.

We will create a graph of "polar sentiment" (nr.positive minus nr.negative) and "positivity index" ($\mbox{positivity index} = \frac{nr.positive}{nr.positive+nr.negative} \times 100$) per chart level. 

Use `pivot_wider()` to separate the sentiments into columns and `mutate()` to create `polarity` (positive - negative)  and a `percent_positive`  (nr.positive/nr.positive+nr∗100
) columns, for a different perspective. 

For the polarity graph, add a y-intercept with `geom_hline()`. Plot the graphs side by side with `grid.arrange()`.

```{r}
prince_polarity_chart <- prince_bing %>%
  count(sentiment, chart_level) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
  mutate(polarity = positive - negative,
    percent_positive = positive / (positive + negative) * 100 )

#Polarity by chart
plot1 <- prince_polarity_chart %>%
  ggplot( aes(chart_level, polarity, fill = chart_level)) +
  geom_col() +
  scale_fill_manual(values = my_colors[3:5]) +
  geom_hline(yintercept = 0, color = "red") +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Polarity By Chart Level")

#Percent positive by chart
plot2 <- prince_polarity_chart %>%
  ggplot( aes(chart_level, percent_positive, fill = chart_level)) +
  geom_col() +
  scale_fill_manual(values = c(my_colors[3:5])) +
  geom_hline(yintercept = 0, color = "red") +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Positivity Index By Chart Level")

grid.arrange(plot1, plot2, ncol = 2)
```


**EXERCISE.** Use the NRC lexicon to repeat the above analysis. Compare the results.

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>

<br>



```{r}
prince_polarity_chart_nrc <- prince_nrc %>%
  filter(sentiment %in% c("positive", "negative")) %>%
  count(sentiment, chart_level) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
  mutate(polarity = positive - negative,
    percent_positive = positive / (positive + negative) * 100 )

#Polarity by chart
plot3 <- prince_polarity_chart_nrc %>%
  ggplot( aes(chart_level, polarity, fill = chart_level)) +
  geom_col() +
  scale_fill_manual(values = my_colors[3:5]) +
  geom_hline(yintercept = 0, color = "red") +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Polarity By Chart Level")

#Percent positive by chart
plot4 <- prince_polarity_chart_nrc %>%
  ggplot( aes(chart_level, percent_positive, fill = chart_level)) +
  geom_col() +
  scale_fill_manual(values = c(my_colors[3:5])) +
  geom_hline(yintercept = 0, color = "red") +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Percent Positive By Chart Level")

grid.arrange(plot3, plot4, ncol = 2)
```

For the NRC lexicon, charted songs have a "neutral" polarity (small). But according to the BING lexicon charted songs show more polarity, highly positive.
The positivity index doesn't seem to vary much bewtween charted and uncharted categories.

If so, what does this tell you about what society wants to hear? Can you even make these assumptions?


# Polarity over time

Since you are looking at sentiment from a polar perspective, you might want to see if it changes over time. 
This time use `geom_smooth()` with the `loess` method for a smoother curve.

Using the BING lexicon:

```{r warning=FALSE}
prince_polarity_year_bing <- prince_bing %>%
  count(sentiment, year) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
  mutate(polarity = positive - negative,
    percent_positive = positive / (positive + negative) * 100)

polarity_over_time_bing <- prince_polarity_year_bing %>%
  ggplot(aes(year, polarity, color = ifelse(polarity >= 0,my_colors[5],my_colors[4]))) +
  geom_col() +
  #geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, aes(color = my_colors[1])) +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Polarity Over Time")

relative_polarity_over_time_bing <- prince_polarity_year_bing %>%
  ggplot(aes(year, percent_positive , color = ifelse(polarity >= 0,my_colors[5],my_colors[4]))) +
  geom_col() +
  #geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, aes(color = my_colors[1])) +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Positivity Index Over Time")

grid.arrange(polarity_over_time_bing, relative_polarity_over_time_bing, ncol = 2)
```

Using the NRC lexicon:

```{r warning=FALSE}
prince_polarity_year_nrc <- prince_nrc %>%
  filter(sentiment %in% c("positive", "negative")) %>%
  count(sentiment, year) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
  mutate(polarity = positive - negative,
    percent_positive = positive / (positive + negative) * 100)

polarity_over_time_nrc <- prince_polarity_year_nrc %>%
  ggplot(aes(year, polarity, color = ifelse(polarity >= 0, my_colors[5], my_colors[4]))) +
  geom_col() +
  #geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, aes(color = my_colors[1])) +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Polarity Over Time")

relative_polarity_over_time_nrc <- prince_polarity_year_nrc %>%
  ggplot(aes(year, percent_positive , color = ifelse(polarity >= 0, my_colors[5], my_colors[4]))) +
  geom_col() +
  #geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, aes(color = my_colors[1])) +
  theme_prince() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Positivity index Over Time")

grid.arrange(polarity_over_time_nrc, relative_polarity_over_time_nrc, ncol = 2)
```

What are your conclusions? Which dictionary would you use for these two indeces?


# Mood by decade cord diagram

We will use the power of the chord diagram to examine the relationships between NRC sentiments and decades. Note that sentiment categories appear on the top part of the ring and decades on the bottom.

```{r}
grid.col = c("1970s" = my_colors[1], "1980s" = my_colors[2], "1990s" = my_colors[3], "2000s" = my_colors[4], "2010s" = "red", "anger" = "grey", "anticipation" = "grey", "disgust" = "grey", "fear" = "grey", "joy" = "grey", "sadness" = "grey", "surprise" = "grey", "trust" = "grey")

decade_mood <-  prince_nrc %>%
  filter(decade != "NA" & !sentiment %in% c("positive", "negative")) %>%
  count(sentiment, decade) %>%
  group_by(decade, sentiment) %>%
  summarise(sentiment_sum = sum(n)) %>%
  ungroup()

decade_mood
```


```{r}
circos.clear()
#Set the gap size
decade_aux <- decade_mood %>% select(decade) %>% n_distinct() #nr of unique values of decade
sentiment_aux <- decade_mood %>% select(sentiment) %>% n_distinct() #nr of unique values of sentiment

circos.par(gap.after = c(rep(5, decade_aux - 1), 10,
                         rep(5, sentiment_aux - 1), 10))
chordDiagram(decade_mood, grid.col = grid.col, transparency = .2)
title("Relationship Between Mood and Decade")
```

Real-Time Sentiment

We will map the analysis of Prince's lyrics to something real over a period of time. 

We have a list of Prince's life events, in the file "data/princeEvents.csv", compiled by Debbie Liske using popular sources such as Rolling Stone Magazine, Biography.com, etc. She selected highly public years that match songs that have release dates in our data set. 

Then use prince_bing and spread() to create a polarity score per year. Join on the events data frame and create a sentiment field so you can fill in colors on your bar chart. As always, use coord_flip() when you're showing large text labels.

```{r message=F}
events <- read_csv("data/princeEvents.csv")
events
```

```{r}
year_polarity_bing <- prince_bing %>%
  group_by(year, sentiment) %>%
  count(year, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n) %>%
  mutate(polarity = positive - negative,
         ratio = polarity / (positive + negative)) #use polarity ratio in next graph

year_polarity_bing
```

```{r}
events %>%
   #Left join gets event years with no releases
   left_join(year_polarity_bing) %>%
   filter(event != " ") %>% 
   mutate(event = reorder(event, year), #Sort chart by desc year
           sentiment = ifelse(positive > negative,
                              "positive", "negative"))
```

```{r}
 events %>%
   #Left join gets event years with no releases
   left_join(year_polarity_bing) %>%
   filter(event != " ") %>% 
   mutate(event = reorder(event, year), #Sort chart by desc year
           sentiment = ifelse(positive > negative,
                              "positive", "negative")) %>%
   ggplot(aes(event, polarity, fill = sentiment)) +
   geom_bar(stat = "identity") +
   theme_minimal() + theme(legend.position = "none") +
   xlab(NULL) +
   ggtitle("Sentiment by Events") +
   coord_flip()
```


# Radar Charts

Another great way to compare sentiment across categories is to use a radar chart, which is also known as a "spider chart". 
You can make this type of charts with the `radarchart` package. 
The charts are useful for seeing which variables have similar values or if there are any outliers for each variable.
You will break this analysis into three different levels: year, chart, and decade. (To save space, I only include code for the specific years.)
Use the `prince_nrc_sub` data set which does not contain the positive and negative sentiments so that the other ones are more visible. 
This time you will first calculate the total count of words by sentiment per year, as well as the total sentiment for the entire year and obtain a percentage (count of sentiment words per year/total per year∗100).

Filter for the specific years 1978, 1994, 1995, and remove the unused variables with `select()`.
Finally, use `pivot_wider()` to create  year columns with percent values in them so that you have one row for each sentiment and a column for each year. 
Then use `chartJSRadar()` to generate an interactive HTML widget. 
You can pass an argument to display data set labels in the mouse over. (FYI, sometimes the J and Y are cropped from the word "joy" by radarchart and it looks like "iov".)


```{r}
#Get the count of words per sentiment per year
year_sentiment_nrc <- prince_nrc_sub %>%
  group_by(year, sentiment) %>%
  count(year, sentiment) %>%
  select(year, sentiment, sentiment_year_count = n)

#Get the total count of sentiment words per year (not distinct)
total_sentiment_year <- prince_nrc_sub %>%
  count(year) %>%
  select(year, year_total = n)

#Join the two and create a percent field
year_radar_chart <- year_sentiment_nrc %>%
  inner_join(total_sentiment_year, by = "year") %>%
  mutate(percent = sentiment_year_count / year_total * 100 ) %>%
  filter(year %in% c("1978", "1994", "1995")) %>%
  select(-sentiment_year_count, -year_total) %>%
  pivot_wider(names_from = year, values_from = percent) %>%
  chartJSRadar(showToolTipLabel = TRUE,
               main = "NRC Years Radar")

year_radar_chart
```

**EXERCISE** produce a sentiment radar chart using the "charted" categories.


# Bigrams Per Decade

So far we have looked at unigrams or single words. We know that this could be taking things out of context as a pre- or proceeding word may give more insight of the sentiment.

We will now look at bigrams, or word pairs.

We need to prepare and clean the data differently now.


The `tidytext` package provides the ability to unnest pairs of words as well as single words. In this case, use `unnest_tokens()` passing the token argument `ngrams`. Since you're just looking at bigrams (two consecutive words), pass `n = 2`. Store the results in `prince_bigrams`.

The `tidyr` package provides the ability to separate the bigrams into individual words using the `separate()` function. 
In order to remove the stop words and undesirable words,  break the bigrams apart and filter out what you don't want, then use `unite()` to put the word pairs back together. This makes it easy to visualize the most common bigrams per decade.

```{r}
prince_bigrams <- prince_data %>%
  unnest_tokens(bigram, lyrics, token = "ngrams", n = 2)
prince_bigrams
```

```{r}
bigrams_separated <- prince_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

undesirable_words <- c("chorus", "repeat", "lyrics",
                       "theres", "bridge", "fe0f", "yeah", "baby",
                       "alright", "wanna", "gonna", "chorus", "verse",
                       "whoa", "gotta", "make", "2",
                       "4", "ohh", "ooh", "uurh", "pheromone", "poompoom", "3121",
                       "matic", " ai ", " ca ", " la ", "hey", " na ",
                       " da ", " uh ", " tin ", "  ll", "transcription",
                       "repeats", "la", "da", "uh", "ah")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%
  filter(!word1 %in% undesirable_words) %>%
  filter(!word2 %in% undesirable_words)

bigrams_filtered
```


```{r}
#Because there is so much repetition in music, also filter out the cases where the two words are the same
#and keep only top 7 bigrams per decade
bigram_decade <- bigrams_filtered %>%
  filter(word1 != word2) %>%
  filter(decade != "NA") %>%
  unite(bigram, word1, word2, sep = " ") %>%
  inner_join(prince_data) %>%
  count(bigram, decade, sort = TRUE) %>%
  group_by(decade) %>%
  slice(seq_len(7)) %>% #keep only top 7 bigrams per decade
  ungroup() %>%
  arrange(decade, n) %>%
  mutate(row = row_number())

bigram_decade
```

Visualise the top 7 bigrams per decade.

```{r}
  bigram_decade %>%
  ggplot(aes(x = row, y = n, fill = decade)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~decade, scales = "free_y") +
  xlab(NULL) + ylab(NULL) +
  scale_x_continuous(  # This handles replacement of row
      breaks = bigram_decade$row, # Notice need to reuse data frame
      labels = bigram_decade$bigram) +
  theme_prince() +
  theme(panel.grid.major.x = element_blank()) +
  ggtitle("Bigrams Per Decade") +
  coord_flip()
```

Using bigrams, you can almost see the common phrases shift from sex, dance and romance to religion and (rainbow) children. 
In case you didn't know, the term "rainbow baby" is sometimes used by parents who are expecting another child after losing a baby to miscarriage. 

# Sentiment with Bigrams

So how do bigrams affect sentiment? This time use the AFINN lexicon to perform sentiment analysis on word pairs, looking at how often sentiment-associated words are preceded by "not" or other negating words.


```{r}
AFINN <- get_sentiments("afinn")

not_words <- bigrams_separated %>%
  filter(word1 == "not") %>%
  inner_join(AFINN, by = c(word2 = "word")) %>%
  count(word2, value, sort = TRUE) %>%
  ungroup()

not_words %>%
  mutate(contribution = n * value) %>%
  arrange(desc(abs(contribution))) %>%
  head(20) %>%
  mutate(word2 = reorder(word2, contribution)) %>%
  ggplot(aes(word2, n * value, fill = n * value > 0)) +
  geom_col(show.legend = FALSE) +
  theme_prince() +
  xlab("Words preceded by \"not\"") +
  ylab("Sentiment value * Number of Occurrences") +
  ggtitle("Sentiment of Words Preceded by Not") +
  coord_flip()
```  
  
Words such as "love", "care", "like" are given a false positive sentiment because the "not" is ignored with single-word analysis. Do the false positive bigrams cancel out the false negative bigrams?
A good topic for further exploration. 

# Network graph to explore bigrams

There are other negation words to consider as well. This time you will create a network graph using the `ggraph` and `igraph` packages. 

We will arrange the words into connected nodes with the negation words at the centers. 
Create the first object from the tidy data set using `graph_from_data_frame()` and then use `ggraph()` to plot it. 
You can highlight the main nodes with a call to `geom_edge_density()`. 
You can get more details of a similar example in Julia Silge and David Robinson's book on Tidy Text Mining.

The package `ggraph` is an extension of `ggplot2` aimed at supporting relational data structures such as networks, graphs, and trees. 
While it builds upon the foundation of `ggplot2` and its API, it comes with its own self-contained set of geoms, facets, etc., as well as adding the concept of layouts to the grammar.

`ggraph` builds upon three core concepts that are quite easy to understand:

* The Layout defines how nodes are placed on the plot, that is, it is a conversion of the relational structure into an x and y value for each node in the graph. `ggraph` has access to all layout functions available in `igraph` and furthermore provides a large selection of its own, such as hive plots, treemaps, and circle packing.

* The Nodes are the connected entities in the relational structure. These can be plotted using the `geom_node_*()` family of geoms. Some node geoms make more sense for certain layouts, e.g. `geom_node_tile()` for treemaps and icicle plots, while others are more general purpose, e.g. `geom_node_point()`.

* The Edges are the connections between the entities in the relational structure. These can be visualized using the `geom_edge_*()` family of geoms that contain a lot of different edge types for different scenarios. Sometimes the edges are implied by the layout (e.g. with treemaps) and need not be plotted, but often some sort of line is warranted.


```{r}
negation_words <- c("not", "no", "never", "without")

negation_bigrams <- bigrams_separated %>%
  filter(word1 %in% negation_words) %>%
  inner_join(AFINN, by = c(word2 = "word")) %>%
  count(word1, word2, value, sort = TRUE) %>%
  mutate(contribution = n * value) %>%
  arrange(desc(abs(contribution))) %>%
  group_by(word1) %>%
  slice(seq_len(20)) %>% # keep the top 20
  arrange(word1, desc(contribution)) %>%
  ungroup()

negation_bigrams
```

```{r}
#Create the network data from a dataframe using the function graph_from_data_frame() from the package igraph
bigram_graph <- negation_bigrams %>%
  graph_from_data_frame()

bigram_graph
```

```{r}
set.seed(123)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(alpha = .25) +
  geom_edge_density(aes(fill = value)) +
  geom_node_point(color = "purple1", size = 1) + #Purple for Prince!
  geom_node_text(aes(label = name),  repel = TRUE) +
  theme_void() + theme(legend.position = "none",
                       plot.title = element_text(hjust = 0.5)) +
  ggtitle("Negation Bigram Network")
```

Here, you can see the word pairs associated with negation words. So if your analysis is based on unigrams and "alone" comes back as negative, the bigram "not alone", as you see above, will have a reverse effect. Some words cross over to multiple nodes which can be seen easily in a visual like this one: for example, "never fair" and "not fair".

# Pairwise Comparisons

Now we will take a look at the correlation between words. 

Which words are most highly associated? 

Use the `pairwise_count()` function from the `widyr` package to identify co-occurrence counts. 
That is, you count the number of times each pair of words appear together within a song. 

syntax:

`pairwise_count(tbl, item, feature)`

`tbl` : the data frame where the data is

`item`: Item to count pairs of; will end up in `item1` and `item2` columns

`feature`: Column within which to count pairs



The `widyr` package takes a tidy data set, and temporarily widens it before returning it to a tidy structure for visualization and further analysis.

To keep it simple, we will focus on four interesting words in Prince's lyrics: love, peace, gangster and hate.


```{r}
pwc <- prince_tidy_filtered %>%
  pairwise_count(word, song, sort = TRUE)
pwc
```
  
```{r}
  pwc <- prince_tidy_filtered %>%
  pairwise_count(word, song, sort = TRUE) %>%
  filter(item1 %in% c("love", "peace", "gangster", "hate")) %>%
  group_by(item1) %>%
  slice(seq_len(7)) %>% # keep top 7
  ungroup() %>%
  mutate(row = -row_number()) #Descending order

pwc %>%
  ggplot(aes(row, n, fill = item1)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  facet_wrap(~item1, scales = "free") +
  scale_x_continuous(  #This handles replacement of row
      breaks = pwc$row, #Notice need to reuse data frame
      labels = pwc$item2) +
  theme_prince() + theme(panel.grid.major.x = element_blank()) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Pairwise Counts") +
  coord_flip()
```  
 
# Pairwise correlation of words presence

Suppose we have two words $X$ and $Y$ and we consider the two binary variables "A: X appears in the document" and "B: Y appears in the document". The binary sample correlation between these two variables, computed from the contingency table formed using all the documents, is called the $\phi$ coefficient.

We can form the following table

|            |   B is true |   B is false  |    Total  |
|:----------:|:-----------:|:-------------:|:---------:|
|A is true   |  $n_{11}$   |   $n_{10}$    |  $n_{1.}$ |
|A is false  |  $n_{01}$   |   $n_{00}$    |  $n_{0.}$ |
|Total       |  $n_{.1}$   |   $n_{.0}$    |    $n$    |


$\phi$ is defined as $\frac{n_{11} n_{00} - n_{10} n_{01}}{\sqrt{n_{0.} n_{.0} n_{1.} n_{.1}}}$ 

When $A$ and $B$ are perfectly associated, that is if $X$ appears, $Y$ appears and if $X$ doesn't appear, $Y$ doesn't appear either, $\phi=1$ because $n_{10} = n_{01} = 0$.

If each time $X$ appears $Y$ doesn't appear, and is when $X$ doesn't appear $Y$ appears, $n_{11} = n_{00} = 0$ and $\phi = -1$.

<br>

The `pairwise_cor()` function in `widyr` gives us the correlation coefficient between words based on how often they appear in the same document.



```{r message=F}
pairw_correlations <- prince_tidy_filtered %>%
  group_by(word) %>%
  filter(n() >= 20) %>%
  pairwise_cor(word, song, sort = TRUE) %>%
  filter(item1 %in% c("gangster", "love", "peace", "hate")) %>%
  group_by(item1) %>%
  top_n(7) %>%
  ungroup()

pairw_correlations
```

```{r}
pairw_correlations %>%
  ggplot(aes(x = reorder_within(item2, correlation, item1), y = correlation, fill = item1)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  scale_x_reordered() +
  facet_wrap(~item1, scales = 'free') +
  theme_prince() + theme(panel.grid.major.x = element_blank()) +
  xlab(NULL) + ylab("Correlation") +
  ggtitle("Pairwise Correlation") +
  coord_flip()
```  
  
After having inspected word co-occurrence and correlation we can go into the next part, topic modelling.



